Coming to Germany: Pre-migration language skills, Job search and Labor market outcomes

### 2018-08-31 Paper 2.tex
Update Paper 2.tex

This paper contributes to two strands of literature, namely, the effects of language skills on labor market outcomes and the job search behavior of immigrants.

@@

Although IV estimations provide a robust estimate of the labor market effect of language skills, they give relatively little insightful information about the mechanism through which such an effect exists (Isphording, 2014). Some authors attempt to document some major channels through which language skills affect the earnings of immigrants, for instance, through complimenting pre--migration schooling and experience (Chiswick and Miller, 2003), occupation level (Berman et al., 2003), increased education (Bleakley and Chin, 2004; Isphording and Sinning, 2012), and occupational choice (Chiswick and Miller, 2010). However, it is empirically not clear whether the pre--migration language skills also work through these channels like contemporaneous language skills.

@@

Like natives, immigrants also use a variety of methods to seek for jobs. These methods include the use of public and private employment agencies, responding to job advertisements on public media (newspapers or the Internet), or relying on their network of family and friends (Battu et al., 2011). Job search theory predicts that people choose their search intensity and search methods by comparing the returns and costs of different options (Holzer, 1988). Hence, as a most important and visible human capital component in the migration history, pre--migration language skills obviously have an impact on the choice of job search methods. And these job search methods, in turn, do matter in subsequent labor market outcomes (Battu et al., 2011).

@@

Cappellari and Tatsiramos (2015) focus on the network quality as a key factor to explain the effectiveness of social contacts on labor market performance. They find that a higher number of employed friends within the closest tie increases both the job finding rate and wages, but only for high-skilled workers. Facchini et al., (2015) use the German Social-Economic Panel (GSOEP) 1996 -- 2011 to find that immigrants who have German friends in their friendship ties tend to have faster cultural assimilation than immigrants who do not. While they argue that labor market integration and education are important determinants of forming friendship ties with natives and not the other way around, there seems to be a strong inter-relation between labor market success and the quality of social networks.

@@

The IAB--SOEP Migration Sample is a household survey undertaken annually by the Institute for Employment Research (IAB) in Nuremberg and the German Socio-Economic Panel (SOEP) at German Institute for Economic Research (DIW Berlin). The first wave of the survey started in 2013 and covered 4,964 persons in 2,723 households. These are immigrants who first appeared in administrative data from the German unemployment insurance in 1995 and are older than 16 years. Some countries of origin are given higher sampling probability to ensure sufficient observations such as immigrants from EU-New Member States and Southern European countries (Bruecker et al., 2014). This new dataset provides a large representative sample of migrants in Germany. With comprehensive information, it also provides new insights to various aspect of immigration such as migration history, education background and labor market performance (Bruecker et al., 2014).

This paper only focuses on immigrants who are from 20 to 65 years of age who are employed on a full--time or part--time basis. I exclude immigrants in education, retirement, civil or military service and self-employment because of their irregular employment activities and unreliable wage information. Further, I restrict the sample to immigrants who first migrated to Germany when they are at least 18 years old, and did not conduct study or vocational training. This restriction is necessary to omit immigrants who are born in Germany, or migrated to Germany as children because the language fluency at the time of migration and at the time of job search could be very different. More importantly, to explore the mediating role of using social networks to obtain jobs on contemporary labor market outcomes, I need the information of job search channels for the current jobs of all individuals. Unfortunately, the IAB-SOEP Migration Sample only provides information about the job search channels of the very first jobs obtained after migration (Dustmann et al., 2016). This information limits the ability to consider immigrants who change their occupations. Therefore, I only consider immigrants who stay in their first jobs after migration so that I am able to follow them from the beginning of searching for jobs until working in the current jobs.

The sample of analysis consists of 965 immigrants in the wage sample and 1,008 immigrants in the level of job complexity sample. I split the whole sample of analysis into two sub-samples to be more transparent due to 43 missing cases of wages. The empirical results shown below are actually similar between these two sub-samples. As I omit immigrants who came to Germany to study or conduct vocational training, there are only three types of immigrants: Labor migrants (30.6 percent), Family reunion (58.7 percent), and humanitarian migrants (10.7 percent). These immigrants mainly come from Eastern Europe (64 percent), 6.5 percent from Southern Europe, approximately 6 percent from Western Europe, more than 6.2 percent from Turkey, and around 17 percent come from the rest of the world. As the IAB-SOEP Migration Sample focuses on the newly arrived immigrants, more than 98 percent people in the final sample came to Germany after 1990, and only 2 percent came during the post-war period (1945) until 1989.

@@

\subsection{Variables}
\textbf{Dependent Variables} \\
I consider two labor market outcome variables in this research. First, the natural logarithm of hourly wages measures the financial rewards to language fluency and the use of social networks to obtain jobs. I convert weekly wages to hourly wages based on the actual number of working hours per week, and deflate the results to the base year in 2010. Second, the level of job complexity is used as a proxy for the quality of jobs. In particular, I extract the last digit as the required levels of qualifications according to the German Classification of Occupations 2010 (\textit{Klassifikation der Berufe - KldB 2010}). There are four levels of job requirement as follows: (1) Unskilled or semi-skilled activities (no vocational qualification required), (2) Specialist activities (at least two years of vocational training), (3) Complex specialist activities (master craftsman or technician with technical school qualification), and (4) Highly complex activities (completed university studies of at least four years).

@@

The mediator variable in this research is the use of social network to obtain the first job after migration to Germany. As only immigrants who stay in their first jobs after migration remain in our sample, this is also the job search method for these immigrants' current jobs. The IAB-SOEP Migration sample includes a number of choices to search for jobs: (1) public employment services, (2) private employment services, (3) job advertisements on public media and (4) personal social networks (friends, family, and business partners from home countries). I generate a dummy variable for using social networks to obtain current jobs which equals to 1 if Yes and 0 otherwise. One immigrant can use multiple job search channels to search for jobs, however, there should be only one realized method that results in the current job. I consider only personal social network because in our sample, more than 65 percent of immigrants indicate that this is their channel of getting current jobs. Due to the limited number of observations, generating a dichotomy indicating the use of social networks to obtain current jobs is better than considering several different job search methods.

@@

To measure the pre-migration language fluency of immigrants in Germany, I use the self-assessed language skills. First, the IAB-SOEP Migration Sample includes three questions to ask about the language skills of immigrants at the very beginning of their migration (the so-called pre-migration language skills) with respect to speaking, writing and reading skills. Immigrants self-assessed these skills on the scale of 1-5 (1: Not all all, 2: Poor, 3: Fair, 4: Good, 5: Very good). I use Principal Component Analysis (PCA) where a single latent component (overall German language skills at migration) is captured by seeking a linear combinations of variables (Speaking, Writing, Reading skills) such that the maximum variance is extracted from these variables (Jolliffe, 2002). Table 1 presents the factor loadings or the correlation coefficients between three original variables and the principal component. All three skills are highly correlated with one another and contribute roughly the same to the single latent factor. The resulting prediction of this PCA is a new composite German language proficiency which captures characteristics of all three individual skills, and is standardized to reflect the z-score. In other words, the new German language proficiency variable has mean 0 and unit standard deviation.

@@

There are four groups of control variables: socio-demographic, human capital, firm characteristics and moderator variables. However, the particular control variables differ in the mediation and the outcome equation. This basically reflects the temporal difference in measuring outcomes of interest. First, in the mediation equation, the dependent variable is the probability of using social network to obtain jobs at the very beginning of the arrival in Germany, therefore, the set of control variables are \textit{gender, age at migration, status at entry, whether they were employed before coming to Germany, and industries}. Second, in the outcome equation, the dependent variable is the earnings and the level of job complexity at the current stage, so the set of control variables are \textit{gender, age, migration period, status at entry, highest qualifications, experience, size of establishments, and industries}.

@@

Third, the average age and age at migration of immigrants is quite high (41 and 32 years old, respectively). More than half of these immigrants were employed before they came to Germany. This is an important piece of information as the job search behavior of the unemployed and the employed is different. Immigrants in the sample mostly achieve at least some forms of school leaving certificates or tertiary degrees, in particular, 23.8 percent are high school graduates, 36.6 percent have vocational training, 33.4 percent are university graduates, and only 6.2 percent do not have any school leaving certificates. In addition, the majority of immigrants works in medium and large establishment with up to 2,000 workers (66 percent). They work mostly in Service sector with more than 60 percent.

>>>>

This paper explores the possible mechanism through which pre--migration language skills influences subsequent labor market outcomes, and therefore, contributes to two strands of literature, namely, the effects of language skills on labor market outcomes and the job search behaviors of immigrants.

@@

Although IV estimations provide a robust estimate of the labor market effect of language skills, they give relatively little insightful information about the mechanism through which such an effect exists (Isphording, 2014). Some authors attempt to document some major channels through which language skills affect the earnings of immigrants, for instance, through complimenting pre--migration schooling and experience (Chiswick and Miller, 2003), occupation level (Berman et al., 2003), increased education (Bleakley and Chin, 2004; Isphording and Sinning, 2012), and occupational choice (Chiswick and Miller, 2010). However, it is empirically not clear whether the pre--migration language skills also work through these channels like contemporaneous language skills and how large the magnitudes of direct and indirect effects are.

@@

Like natives, immigrants also use a variety of methods to seek for jobs. These methods include the use of public and private employment agencies, responding to job advertisements on public media (newspapers or the Internet), or relying on their network of family and friends (Battu et al., 2011). Job search theory predicts that people choose their search intensity and search methods by comparing the returns and costs of different options (Holzer, 1988). Chiswick and Miller (1995) also argue that a higher level of language proficiency facilitates consumption and reduces search costs. Hence, as the most important and visible human capital component in the migration history, pre--migration language skills presumably have a strong impact on the choice of job search methods. And these job search methods, in turn, do matter in subsequent labor market outcomes (Battu et al., 2011).

@@

Cappellari and Tatsiramos (2015) focus on the network quality as a key factor to explain the effectiveness of social contacts on labor market performance. They find that a higher number of employed friends within the closest tie increases both the job finding rate and wages, but only for high-skilled workers. Facchini et al., (2015) use the German Social-Economic Panel (GSOEP) 1996 -- 2011 to find that immigrants who have German friends in their friendship ties tend to have faster cultural assimilation than immigrants who do not. While they argue that labor market integration and education are important determinants of forming friendship ties with natives and not the other way around, there seems to be a strong inter-relation between the quality of social networks and the labor market successes.

@@

The IAB--SOEP Migration Sample is a household survey undertaken annually by the Institute for Employment Research (IAB) in Nuremberg and the German Socio-Economic Panel (SOEP) at German Institute for Economic Research (DIW Berlin). The first wave of the survey started in 2013 and covered 4,964 persons in 2,723 households. These are immigrants who first appeared in the administrative data from the German unemployment insurance in 1995 and are older than 16 years. Some countries of origin are given higher sampling probability to ensure sufficient observations such as immigrants from EU-New Member States and Southern European countries (Bruecker et al., 2014). This new dataset provides a large representative sample of migrants in Germany. With comprehensive information, it also provides new insights to various aspect of immigration such as migration history, education background and labor market performance (Bruecker et al., 2014).

This paper only focuses on immigrants who are from 20 to 65 years of age and are employed on a full--time or part--time basis. I exclude immigrants in education, retirement, civil or military service, mini--jobs and self--employment because of their irregular employment activities and unreliable wage information. Further, I restrict the sample to immigrants who first migrated to Germany when they are at least 18 years old, and with any purposes other than to conduct study or vocational training. This restriction is necessary to omit immigrants who are born in Germany, or migrated to Germany as children or to conduct long--term trainings because the language fluency at the time of migration and at the time of job search could be very different. More importantly, to explore the mediating role of using social networks to obtain jobs on contemporary labor market outcomes, I need the information of job search channels for the current jobs of all individuals. Unfortunately, the IAB-SOEP Migration Sample only provides information about the job search channels of the very first jobs obtained after migration (Dustmann et al., 2016). This information limits the ability to consider immigrants who change their occupations. Therefore, I only consider immigrants who stay in their first jobs after migration so that I am able to follow them from the beginning of searching for jobs until working in the current jobs.

The sample of analysis consists of 965 immigrants in the wage sample and 1,008 immigrants in the level of job complexity sample. I split the whole sample of analysis into two sub-samples to be more transparent due to 43 missing cases of wages. The empirical results shown below are actually similar between these two sub-samples. As I omit immigrants who came to Germany to study or conduct vocational training, there are only three types of immigrants: Labor migrants (30.6 percent), Family reunion (58.7 percent), and humanitarian migrants (10.7 percent). These immigrants mainly come from Eastern Europe (64 percent), the rest are from Southern Europe (6.5 percent), Western Europe (approximately 6 percent), Turkey (6.2 percent), and the rest of the world (17 percent). As the IAB--SOEP Migration Sample focuses on the newly arrived immigrants, more than 98 percent people in the final sample came to Germany after 1990, and only 2 percent came during the post-war period (1945) until 1989.

@@

\subsection{Variables}
\textbf{Dependent Variables} \\
I consider two labor market outcome variables in this research. First, the natural logarithm of hourly wages measures the financial rewards to language fluency and the use of social networks to obtain jobs. I convert weekly wages to hourly wages based on the actual number of working hours per week, and deflate the results to the base year in 2010. Second, the level of job complexity is used as a proxy for the quality of jobs. In particular, I extract the last digit in the occupational codes of the German Classification of Occupations 2010 (\textit{Klassifikation der Berufe - KldB 2010}) as the required levels of qualifications. There are four levels of job requirements as follows: (1) Unskilled or semi-skilled activities (no vocational qualification required), (2) Specialist activities (at least two years of vocational training), (3) Complex specialist activities (master craftsman or technician with technical school qualification), and (4) Highly complex activities (completed university studies of at least four years).

@@

The mediator variable in this research is the use of social network to obtain the first job after migration to Germany. As only immigrants who stay in their first jobs after migration remain in our sample, this is also the job search method of the current jobs of these immigrants. The IAB-SOEP Migration sample includes a number of choices to search for jobs: (1) public employment services, (2) private employment services, (3) job advertisements on public media and (4) personal social networks (friends, family, and business partners from home countries). I generate a dummy variable for using social networks to obtain current jobs which equals to 1 if Yes and 0 otherwise. One immigrant can use multiple job search channels to search for jobs, however, there should be only one realized method that results in the current job. I only consider two job search options, either using personal social networks or not, because in our sample, more than 65 percent of immigrants indicate that this is their channel of getting current jobs. Due to the limited number of observations, generating a dichotomy indicating the use of social networks to obtain current jobs is better than considering several different job search methods.

@@

To measure the pre-migration language fluency of immigrants in Germany, I use the self--assessed language skills. First, the IAB-SOEP Migration Sample includes three questions to ask about the language skills of immigrants at the very beginning of their migration (the so-called pre--migration language skills) with respect to speaking, writing and reading skills. Immigrants self--assessed these skills on the scale of 1-5 (1: Not all all, 2: Poor, 3: Fair, 4: Good, 5: Very good). I use Principal Component Analysis (PCA) where a single latent component (overall German language skills at migration) is captured by seeking a linear combinations of variables (Speaking, Writing, Reading skills) such that the maximum variance is extracted from these variables (Jolliffe, 2002). Table 1 presents the factor loadings or the correlation coefficients between three original variables and the principal component. All three skills are highly correlated with one another and contribute roughly the same to the single latent factor. The resulting prediction of this PCA is a new composite German language proficiency which captures characteristics of all three individual skills, and is standardized to reflect the z-scores. In other words, the new German language proficiency variable has mean 0 and unit standard deviation.

@@

There are four groups of control variables: socio-demographic, human capital, firm characteristics and moderator variables. However, the particular control variables differ in the mediation and the outcome equation. This basically reflects the temporal difference in measuring outcomes of interest. First, in the mediation equation, the dependent variable is the probability of using social network to obtain jobs at the very beginning of the arrival in Germany, therefore, the set of control variables are \textit{gender, age at migration, status at entry, whether they were employed before coming to Germany, and industries}. Second, in the outcome equation, the dependent variable is the earnings and the level of job complexity at the current stage, so the set of control variables are \textit{gender, age, migration period, status at entry, highest qualifications, experience, size of establishments, industries and unemployment rates at the Federal State level}.

@@

Third, the average age and age at migration of immigrants is quite high (41 and 32 years old, respectively). More than half of these immigrants were employed before they came to Germany. This is an important piece of information as the job search behaviors of the unemployed and the employed is different. Immigrants in the sample mostly achieve at least some forms of school leaving certificates or tertiary degrees, in particular, 23.8 percent are high school graduates, 36.6 percent have vocational training, 33.4 percent are university graduates, and only 6.2 percent do not have any school leaving certificates. In addition, the majority of immigrants works in medium and large establishment with up to 2,000 workers (66 percent). They work mostly in Service sector with more than 60 percent.
